---
title: "p53_kd_DNA_methyl_analysis"
output: html_document
date: "2025-12-19"
editor_options:
  chunk_output_type: console
---

## Setup

```{r paths}
# ---- User-configurable paths ----
# Update these to point at your own local copies of the data before running.
project_dir <- "/Users/wakaitly/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Wang, Richard's files - Angiosarcoma project/Manuscript/Methylation array/KW_DNA_methylation_analysis/"
idat_dir <- file.path(project_dir, "p53KD_idat_files")
sample_sheet_pattern <- "Wang_16_samplesheet_12082025_csv_p53KD_only.csv"
beta_gene_csv <- "beta_values_table_funnorm_ALL_samples_gene_info.csv"
minfi_beta_csv <- "beta_values_table_from_minfi_p53KD.csv"
output_dir <- file.path(project_dir, "output_plots")

setwd(project_dir)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

```{r setup}
# Package installation (run once, then leave commented out):
# BiocManager::install(c("limma", "minfi", "IlluminaHumanMethylationEPICv2anno.20a1.hg38", "IlluminaHumanMethylationEPICv2manifest"))
# install.packages(c("RColorBrewer", "tidyverse", "VennDiagram", "pheatmap"))

library("limma")
library("minfi")
library("RColorBrewer")
library("tidyverse")
library("VennDiagram")
library("pheatmap")
library("IlluminaHumanMethylationEPICv2manifest") #just the annotation parts from missMethyl
library("IlluminaHumanMethylationEPICv2anno.20a1.hg38")

EPIC_ann <- getAnnotation(IlluminaHumanMethylationEPICv2anno.20a1.hg38)
```

## Load sample sheet

```{r make samples list}
list.files(idat_dir, recursive = TRUE)

targets <- read.metharray.sheet(project_dir, pattern = sample_sheet_pattern)

# Note: read.metharray.sheet only works correctly with a single idat folder
# present in the working directory - move any extra idat folders up one
# level when they're not part of the current run.
```

## Build and preprocess the methylation data

```{r rebuild data}
rgSet <- read.metharray.exp(targets = targets)

MSet <- preprocessRaw(rgSet) # convert to MethylSet format
ratioSet <- ratioConvert(MSet, what = "both", keepCN = TRUE)
gset <- mapToGenome(ratioSet)

# pull out beta values and save to save local disk space
beta <- getBeta(gset)
write_delim(rownames_to_column(as.data.frame(beta), "probe_ID"), minfi_beta_csv, delim = ",", col_names = TRUE)

# normalize
mSetSq <- preprocessFunnorm(rgSet)

# filter out probes that fail detection in any sample
detP <- detectionP(rgSet)
detP <- detP[match(featureNames(mSetSq), rownames(detP)), ]
keep <- rowSums(detP < 0.01) == ncol(mSetSq)
mSetSqFlt <- mSetSq[keep, ]
```

## Sample clustering (MDS)

```{r MDS plot}
custom_pal <- c("forestgreen", "red", "firebrick4", "blue", "cyan3")

png(file.path(output_dir, "MDS_plot.png"), width = 1800, height = 1200, res = 200)
layout(matrix(1:2, nrow = 1), widths = c(3, 1))

par(mar = c(4.1, 4.1, 4.1, 1))
plotMDS(getM(mSetSq), top = 1000, gene.selection = "common", pch = 16, col = custom_pal[factor(targets$Sample_Name)])

par(mar = c(4.1, 0, 4.1, 0))
plot.new()
legend("left", legend = levels(factor(targets$Sample_Name)), text.col = custom_pal, cex = 0.8, bty = "n")
dev.off()
```

## Differential methylation analysis

QC of the raw beta-value distributions showed that the large majority of probes are strongly bimodal - most CpGs sit near fully unmethylated (β < 0.2) or fully methylated (β > 0.8), typical of EPIC array data. Fold-change on beta values was not informative on this bounded 0-1 scale (ratios cluster tightly around 1 regardless of true signal), so CpGs were instead classified as differentially methylated between conditions using an absolute difference threshold of |Δβ| ≥ 0.2.

```{r Diff exp: threshold-based differential methylation}
beta_gene <- read.csv(beta_gene_csv) #generated in DNAme_analysis_clean.Rmd

# differentially methylated CpGs per comparison (|delta beta| >= 0.2)
diff_CpG_R132C_v_ctl <- filter(beta_gene, abs(p53KD_R132C_betas - p53KD_ctl_betas) >= 0.2)
diff_CpG_R132H_v_ctl <- filter(beta_gene, abs(p53KD_R132H_betas - p53KD_ctl_betas) >= 0.2)
diff_CpG_R132C_v_Ivo <- filter(beta_gene, abs(p53KD_R132C_betas - p53KD_R132C_Ivo_betas) >= 0.2)
diff_CpG_R132H_v_Ivo <- filter(beta_gene, abs(p53KD_R132H_betas - p53KD_R132H_Ivo_betas) >= 0.2)

nrow(diff_CpG_R132C_v_ctl) # 683
nrow(diff_CpG_R132H_v_ctl) # 533
nrow(diff_CpG_R132C_v_Ivo) # 142
nrow(diff_CpG_R132H_v_Ivo) # 123
```

### Overlap between R132H and R132C (mutant vs. parental)

```{r venn diagram}
venn.diagram(x = list(diff_CpG_R132H_v_ctl$probeID, diff_CpG_R132C_v_ctl$probeID),
             category.names = c("R132H", "R132C"),
             filename = file.path(output_dir, "p53_H_v_C_venn.png"),
             imagetype = "png")
# 254 CpGs overlap between the R132H and R132C mutant-vs-parental sets
# (~37% of R132C's 683, ~48% of R132H's 533)
```

### Heatmaps

```{r heatmaps}
p53kd_cols <- c("p53KD_ctl_betas", "p53KD_R132H_betas", "p53KD_R132H_Ivo_betas", "p53KD_R132C_betas", "p53KD_R132C_Ivo_betas")

# all comparisons combined
diff_CpG_inclusive <- filter(beta_gene,
              abs(p53KD_R132C_betas - p53KD_ctl_betas) >= 0.2 |
              abs(p53KD_R132H_betas - p53KD_ctl_betas) >= 0.2 |
              abs(p53KD_R132C_betas - p53KD_R132C_Ivo_betas) >= 0.2 |
              abs(p53KD_R132H_betas - p53KD_R132H_Ivo_betas) >= 0.2
              ) # 1137 CpGs
pheatmap(t(diff_CpG_inclusive[p53kd_cols]), show_colnames = FALSE,
         filename = file.path(output_dir, "heatmap_diff_CpG_C_and_H_p53KD.png"))

# R132C only (vs ctl or vs Ivo)
diff_CpG_C <- filter(beta_gene,
              abs(p53KD_R132C_betas - p53KD_ctl_betas) >= 0.2 |
              abs(p53KD_R132C_betas - p53KD_R132C_Ivo_betas) >= 0.2
              ) # 773 CpGs
pheatmap(t(diff_CpG_C[c("p53KD_ctl_betas", "p53KD_R132C_betas", "p53KD_R132C_Ivo_betas")]), show_colnames = FALSE,
         filename = file.path(output_dir, "heatmap_diff_CpG_C_only_p53KD.png"))

# R132H only (vs ctl or vs Ivo)
diff_CpG_H <- filter(beta_gene,
              abs(p53KD_R132H_betas - p53KD_ctl_betas) >= 0.2 |
              abs(p53KD_R132H_betas - p53KD_R132H_Ivo_betas) >= 0.2
              ) # 635 CpGs
pheatmap(t(diff_CpG_H[c("p53KD_ctl_betas", "p53KD_R132H_betas", "p53KD_R132H_Ivo_betas")]), show_colnames = FALSE,
         filename = file.path(output_dir, "heatmap_diff_CpG_H_only_p53KD.png"))

# R132C vs. parental only
pheatmap(t(diff_CpG_R132C_v_ctl[c("p53KD_ctl_betas", "p53KD_R132C_betas")]), show_colnames = FALSE,
         filename = file.path(output_dir, "heatmap_diff_CpG_C_only_mut_v_parent_p53KD.png"))

# R132H vs. parental only
pheatmap(t(diff_CpG_R132H_v_ctl[c("p53KD_ctl_betas", "p53KD_R132H_betas")]), show_colnames = FALSE,
         filename = file.path(output_dir, "heatmap_diff_CpG_H_only_mut_v_parent_p53KD.png"))
```
